mathml: Translate R expressions to MathML and LaTeX/MathJax

### 2022-12-05 Draft2.Rmd
so many changes

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(plotly)
library(ggplot2)
library(palmerpenguins)

catmath <- function(term)
{
  s <- substitute(term)
  c <- canonical(s)
  mathml(c)
}

catmathjax <- function(term)
{
  s <- substitute(term)
  c <- canonical(s)
  mathjax(c)
}

@@

```{r, echo=TRUE, results='asis', eval=knitr::is_html_output()}
library(rolog)
library(mathml)
term = quote(pbinom(k, N, p))

# Pretty print
mathml(term)
```

```{r, echo=TRUE, results='asis', eval=knitr::is_latex_output()}
library(rolog)
library(mathml)
term = quote(pbinom(k, N, p))

# Pretty print
mathjax(term)
```

@@

```{r,  results = 'asis', echo = TRUE, eval=knitr::is_html_output()}
term = quote(integrate(sin, 0L, 2L*pi))
mathml(term)
```

@@

```{r, echo = TRUE}
library(rolog)

# A custom function
g <- function(u)
{
  sin(u)
}
# # Mixture of (partially) named and positional arguments in unusual order
term <- quote(2L * integrate(low=-Inf, up=Inf, g)$value)
# mathml(canonical(term))
# # It is a bit of a mystery that R knows the result of this integral.
eval(term)
```

@@

<!-- Some packages on interactive graphics include \CRANpkg{plotly} [@plotly] that interfaces with Javascript for web-based interactive graphics, \CRANpkg{crosstalk} [@crosstalk] that specializes cross-linking elements across individual graphics.  The recent R Journal paper \CRANpkg{tsibbletalk} [@RJ-2021-050] provides a good example of including interactive graphics into an article for the journal. It has both a set of linked plots, and also an animated gif example, illustrating linking between time series plots and feature summaries. -->

@@

```{r, results="asis", echo=TRUE,  eval=knitr::is_html_output()}
catmath(x + y)
catmathjax(x + y)

catmath(x - y)
catmathjax(x - y)

catmath(x %+-% y)
catmathjax(x %+-% y)

catmath(x*y)
catmathjax(x*y)

catmath(x %*% y)
catmathjax(x %*% y)

catmath(x %.% y)
catmathjax(x %.% y)

catmath(X %x% Y)
catmathjax(X %x% Y)

catmath(x/y)
catmathjax(x/y)

catmath(x %/% y)
catmathjax(x %/% y)
```

@@

```{r, results="asis", echo=TRUE,  eval=knitr::is_html_output()}
catmath(a["x"])
catmathjax(a["x"])

catmath(a^x)
catmathjax(a^x)

catmath(x["b"]^"a")
catmathjax(x["b"]^"a")

@@

```{r, results="asis", echo=TRUE,  eval=knitr::is_html_output()}
catmath(frac(x, y))
catmathjax(frac(x, y))

catmath(over(x, y))
catmathjax(over(x, y))

catmath(choose(n, k))
catmathjax(choose(n, k))

catmath(lchoose(n, k))
catmathjax(lchoose(n, k))
```

@@

```{r, results="asis", echo=TRUE,  eval=knitr::is_html_output()}

catmath(sqrt(x))
catmathjax(sqrt(x))

@@

```{r, results="asis", echo=TRUE,  eval=knitr::is_html_output()}
catmath(x==y)
catmathjax(x==y)

catmath(x < y)
catmathjax(x < y)

catmath(x <= y)
catmathjax(x <= y)

catmath(x >= y)
catmathjax(x >= y)

catmath(!x)
catmathjax(!x)

catmath(x != y)
catmathjax(x != y)

catmath(X ~ Y)
catmathjax(X ~ Y)

catmath(x %~~% y)
catmathjax(x %~~% y)

catmath(x %==% y)
catmathjax(x %==% y)

catmath(x %=~% y)
catmathjax(x %=~% y)

catmath(x %prop% y)
catmathjax(x %prop% y)

@@

```{r, results="asis", echo=TRUE,  eval=knitr::is_html_output()}
catmath(a %in% A)
catmathjax(a %in% A)

catmath(intersect(A, B))
catmathjax(intersect(A, B))

catmath(union(A, B))
catmathjax(union(A, B))

catmath(is.null(intersect(A, B)))
catmathjax(is.null(intersect(A, B)))
```

@@

```{r, results="asis", echo=TRUE,  eval=knitr::is_html_output()}
# x %<->% y
catmath(x %<->% y)
catmathjax(x %<->% y)

# x %->% y
catmath(x %->% y)
catmathjax(x %->% y)

# x %<-% y
catmath(x %<-% y)
catmathjax(x %<-% y)

# x %up% y:
catmath(x %up% y)
catmathjax(x %up% y)

# x %down% y:
catmath(x %down% y)
catmathjax(x %down% y)

# x %<=>% y: solved it using iff
catmath(x %<=>% y)
catmathjax(x %<=>% y)

# x %=>% y: solved using rArr
catmath(x %=>% y)
catmathjax(x %=>% y)

# x %<=% y: solved using lArr
catmath(x %<=% y)
catmathjax(x %<=% y)

# x %dblup% y:
catmath(x %dblup% y)
catmathjax(x %dblup% y)

# x %dbldown% y:
catmath(x %dbldown% y)
catmathjax(x %dbldown% y)

@@

```{r, results="asis", echo=TRUE,  eval=knitr::is_html_output()}
catmath(hat(x))
catmathjax(hat(x))

catmath(hat(xy))
catmathjax(hat(xy))

catmath(mean(x))
catmathjax(mean(x))

catmath(mean(xy))
catmathjax(mean(xy))

@@

```{r, results="asis", echo=TRUE,  eval=knitr::is_html_output()}

catmath(Alpha-Omega)
catmathjax(Alpha-Omega)

catmath(alpha-omega)
catmathjax(alpha-omega)

catmath(Inf)
catmathjax(Inf)

## Operators
A list of possible operators is available here.

```{r, results='asis', echo=TRUE,  eval=knitr::is_html_output()}

catmath(sum(x[i], i==1L, n))
catmathjax(sum(x[i], i==1L, n))

# fixme: braces not needed when right to *
catmath(a * sum(b))
catmathjax(a * sum(b))

# Multiplication (capital Pi) is missing

catmath(integrate(sin, 0L, 2L*pi))
catmathjax(integrate(sin, 0L, 2L*pi))

catmath(A & B)
catmathjax(A & B)

catmath(A | B)
catmathjax(A | B)

catmath(!A)
catmathjax(!A)

catmath(xor(A, B))
catmathjax(xor(A, B))

catmath(exp(2L*pi*i))
catmathjax(exp(2L*pi*i))

catmath(expm1(2L*pi*i))
catmathjax(expm1(2L*pi*i))

catmath(log(x))
catmathjax(log(x))

catmath(log10(x))
catmathjax(log10(x))

catmath(log2(x))
catmathjax(log2(x))

catmath(logb(x, e))
catmathjax(logb(x, e))

catmath(log1p(x))
catmathjax(log1p(x))

catmath(ceiling(pi))
catmathjax(ceiling(pi))

catmath(floor(pi))
catmathjax(floor(pi))

catmath(sin(pi/2L))
catmathjax(sin(pi/2L))

catmath(cos(pi/2L))
catmathjax(cos(pi/2L))

catmath(tan(pi/2L))
catmathjax(tan(pi/2L))

@@

```{r, results='asis', echo=TRUE,  eval=knitr::is_html_output()}
catmath(x^y + z)
catmathjax(x^y + z)

catmath(x^(y + z))
catmathjax(x^(y + z))

# without parenthesis
mathml(canonical(call("^", as.symbol("x"), call("+", as.symbol("y"), as.symbol("z")))))
mathjax(canonical(call("^", as.symbol("x"), call("+", as.symbol("y"), as.symbol("z")))))

# shortcut
catmath(x^{y + z})
catmathjax(x^{y + z})

@@

```{r, results='asis', echo=TRUE,  eval=knitr::is_html_output()}
catmath(abs(x))
catmathjax(abs(alpha))

@@

```{r, results='asis', echo=TRUE,  eval=knitr::is_html_output()}

vec <- 1:3
mathml(canonical(vec))
mathml(vec)
mathjax(canonical(vec))
mathjax(vec)

mat <- matrix(data = c(2L, 3L, 4L, 5L), nrow = 2, ncol = 2)
mathml(canonical(mat))
mathjax(canonical(mat))

@@

```{r, results="asis", echo=FALSE,  eval=knitr::is_html_output()}

# Improve display
hook(quote(mu_A), quote(mu["A"]))
hook(quote(mu_B), quote(mu["B"]))
hook(quote(sigma_A), quote(sigma["A"]))
hook(quote(sigma_B), quote(sigma["B"]))
hook(quote(mu_M), quote(mu["M"]))
hook(quote(zn), quote(z[tau]))
hook(quote(M), quote(overline(X)))
```

@@

```{r, results="asis", echo=TRUE,  eval=knitr::is_html_output()}
f <- function(tau, c, mu_A, sigma_A, mu_B, sigma_B)
{ dfrac(c, mu_A) + (dfrac(1L, mu_A) - dfrac(1L, mu_A + mu_B)) *
    ((mu_A*tau - c) * pnorm(dfrac(c - mu_A*tau, sqrt(sigma_A^2L*tau)))
      - (mu_A*tau + c) * exp(dfrac(2L*mu_A*c, sigma_A^2L))
        * pnorm(dfrac(-c - mu_A*tau, sqrt(sigma_A^2L*tau))))
}

mathml(f)
```

@@

````{r, results="asis", echo=FALSE,  eval=knitr::is_html_output()}
mu_A = 0.53
mu_B = 1.32
sigma_A = 4.4
sigma_B = 18.3
tau = c(-Inf, -167, -133, -100, -67, -33, 0, 33, 67, 100, 133, 167, Inf)
c = 100
mu_M = 161
````

```{r, results="asis", echo=FALSE,  eval=knitr::is_html_output()}

# Conditions

@@

mathml(g)

@@

```{r, results="asis", echo=FALSE,  eval=knitr::is_html_output()}
mu_A = 0.53
mu_B = 1.32
sigma_A = 4.4
sigma_B = 18.3
tau = -167
c = 100
mu_M = 161

g(tau, c, mu_A, sigma_A, mu_B, sigma_B)

@@

```{r, results="asis", echo=FALSE,  eval=knitr::is_html_output()}
g(tau, c, mu_A, sigma_A, mu_B, sigma_B) + mu_M
```

@@

```{r, results="asis", echo=FALSE,  eval=knitr::is_html_output()}

@@

h <- function(tau, c, mu_A, sigma_A, mu_B, sigma_B, mu_M)
{
  G(tau, c, mu_A, sigma_A, mu_B, sigma_B) + mu_M
}
mathml(h)
```

```{r, results="asis", echo=FALSE,  eval=knitr::is_html_output()}

@@

#h(tau, c, mu_A, sigma_A, mu_B, sigma_B, mu_M)

@@

```{r, results="asis", echo=FALSE,  eval=knitr::is_html_output()}

zf <-  function(M, s, N, tau, c, mu_A, sigma_A, mu_B, sigma_B, mu_M)
{
  E <- h(tau, c, mu_A, sigma_A, mu_B, sigma_B, mu_M)
  dfrac(M-E, s/sqrt(N))
}

mathml(zf)
```

```{r, results="asis", echo=FALSE,  eval=knitr::is_html_output()}

# The vector X contains the observed values, the vector s their observed standard deviation and N = 400 are the number of trials.

@@

zn <- zf(M, s, N, tau, c, mu_A, sigma_A, mu_B, sigma_B, mu_M)
# zn

@@

```{r, results="asis", echo=FALSE,  eval=knitr::is_html_output()}

par <- c(mu_A = 0.53, sigma_A = 4.4, mu_B = 1.32, sigma_B = 18.3, mu_M = 161)

@@

mathml(gof)

n <- gof(par, tau, M, s, N)
#n

#pchisq(n, 8, lower.tail = FALSE, log.p = FALSE)

```{r, results="asis", echo=FALSE,  eval=knitr::is_html_output()}

lsq = function(tau, M, s, N)
{
  start = c(mu_A = 0.3, sigma_A = 2, mu_B = 1 , sigma_B = 5, mu_M = 200)
  optim(start, fn = gof,
    tau = tau, M = M, s = s, N = N, hessian = TRUE, control = list(fnscale = 1))
}

# mathml(lsq)

@@

````{r, results="asis", echo=TRUE,  eval=knitr::is_html_output()}
f <- function(tau)
{ dfrac(c, mu["A"]) + (dfrac(1L, mu["A"]) - dfrac(1L, mu["A"] + mu["B"]) * 
    ((mu["A"]*tau - c) * pnorm(dfrac(c - mu["A"]*tau, sqrt(sigma["A"]^2L*tau)))
      - (mu["A"]*tau + c) * exp(dfrac(2L*mu["A"]*tau, sigma["A"]^2L))
        * pnorm(dfrac(-c - mu["A"]*tau, sqrt(sigma["A"]^2L*tau)))))
}

attributes(f)$name = quote(f)
mathml(f)
````

@@

````{r, results="asis", echo=TRUE,  eval=knitr::is_html_output()}
f <- function(tau)
{ dfrac(c, mu["A"]) + (dfrac(1L, mu["A"]) - dfrac(1L, mu["A"] + mu["B"])) * 
    ((mu["A"]*tau - c) * pnorm(dfrac(c - mu["A"]*tau, sqrt(sigma["A"]^2L*tau)))
      - (mu["A"]*tau + c) * exp(dfrac(2L*mu["A"]*tau, sigma["A"]^2L))
        * pnorm(dfrac(-c - mu["A"]*tau, sqrt(sigma["A"]^2L*tau))))
}

attributes(f)$name = quote(f)
mathml(f)
````

>>>>

```{r setup, include=FALSE}
# knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(plotly)
library(ggplot2)
library(palmerpenguins)

mathout <- function(term)
{
  if(knitr::is_html_output())
    return(mathml(term))

  if(knitr::is_latex_output())
    return(mathjax(term))

  warning("no output specified")  
  mathjax(term)
}

catmath <- function(term)
{
  s <- substitute(term)
  mathout(canonical(s))
}
```

@@

```{r, echo=TRUE, results='asis'}
library(rolog)
library(mathml)
term = quote(pbinom(k, N, p))

# Pretty print
mathout(term)
```

@@

```{r,  results = 'asis', echo = TRUE}
term = quote(integrate(sin, 0L, 2L*pi))
mathout(term)
```

@@

```{r, echo = TRUE, results='asis'}
# A custom function
g <- function(u)
{
  sin(u)
}

# Mixture of (partially) named and positional arguments in unusual order
term <- call("integrate", low=-Inf, up=Inf, g)
# mathout(canonical(term))

# term <- quote(2L * integrate(low=-Inf, up=Inf, g)$value)
# mathout(canonical(term))

# It is a bit of a mystery that R knows the result of this integral.
# eval(term)
```

@@

@@

```{r, results="asis", echo=TRUE}
catmath(x + y)

catmath(x - y)

# Note wrong precedence of custom operator
catmath(mean(m) %+-% 1.96 * s/sqrt(N))

# This is what we actually want
catmath(mean(m) %+-% (1.96 * s/sqrt(N)))

# Without parentheses
catmath(mean(m) %+-% {1.96 * s/sqrt(N)})

catmath(x*y)

catmath(x %*% y)

catmath(x %.% y)

catmath(X %x% Y)

catmath(x/y)

catmath(x %/% y)
```

@@

```{r, results="asis", echo=TRUE}
catmath(a["x"])

catmath(a^x)

catmath(x["b"]^"a")
```

@@

```{r, results="asis", echo=TRUE}
catmath(frac(x, y))

catmath(over(x, y))

catmath(choose(n, k))

catmath(lchoose(n, k))
```

@@

```{r, results="asis", echo=TRUE}
catmath(sqrt(x))
```

@@

```{r, results="asis", echo=TRUE}
catmath(x==y)

catmath(x < y)

catmath(x <= y)

catmath(x >= y)

catmath(!x)

catmath(x != y)

catmath(X ~ Y)

catmath(x %~~% y)

catmath(x %==% y)

catmath(x %=~% y)

catmath(x %prop% y)
```

@@

```{r, results="asis", echo=TRUE}
catmath(a %in% A)

catmath(intersect(A, B))

catmath(union(A, B))

catmath(is.null(intersect(A, B)))
```

@@

```{r, results="asis", echo=TRUE}
# x %<->% y
catmath(x %<->% y)

# x %->% y
catmath(x %->% y)

# x %<-% y
catmath(x %<-% y)

# x %up% y:
catmath(x %up% y)

# x %down% y:
catmath(x %down% y)

# x %<=>% y: solved it using iff
catmath(x %<=>% y)

# x %=>% y: solved using rArr
catmath(x %=>% y)

# x %<=% y: solved using lArr
catmath(x %<=% y)

# x %dblup% y:
catmath(x %dblup% y)

# x %dbldown% y:
catmath(x %dbldown% y)
```

@@

```{r, results="asis", echo=TRUE}
catmath(hat(x))

catmath(hat(xy))

catmath(mean(x))

catmath(mean(xy))
```

@@

```{r, results="asis", echo=TRUE}
catmath(Alpha-Omega)

catmath(alpha-omega)

catmath(pi-Pi)

catmath(Inf)
```

## Operators

A list of possible operators is available here.

```{r, results='asis', echo=TRUE}
catmath(sum(x[i], i==1L, n))

# fixme: braces not needed when right to *
catmath(a * sum(b))

# Multiplication (capital Pi) is missing
catmath(a * prod(b))

catmath(integrate(sin, 0L, 2L*pi))

catmath(A & B)

catmath(A | B)

catmath(!A)

catmath(xor(A, B))

catmath(exp(2L*pi*i))

catmath(expm1(2L*pi*i))

catmath(log(x))

catmath(log10(x))

catmath(log2(x))

catmath(logb(x, e))

catmath(log1p(x))

catmath(ceiling(pi))

catmath(floor(pi))

catmath(sin(pi/2L))

catmath(cos(pi/2L))

catmath(tan(pi/2L))
```

@@

```{r, results='asis', echo=TRUE}
catmath(x^y + z)

catmath(x^(y + z))

# without parenthesis
mathout(canonical(call("^", as.symbol("x"), call("+", as.symbol("y"), as.symbol("z")))))

# shortcut
catmath(x^{y + z})
```

@@

```{r, results='asis', echo=TRUE}
catmath(abs(x))
```

@@

```{r, results='asis', echo=TRUE}
vec <- 1:3
mathout(canonical(vec))
mathout(vec)

mat <- matrix(data = c(2L, 3L, 4L, 5L), nrow = 2, ncol = 2)
mathout(canonical(mat))
```

@@

```{r, echo=FALSE}
# Improve display
hook(quote(mu_A), quote(mu["A"]))
hook(quote(mu_B), quote(mu["B"]))
hook(quote(sigma_A), quote(sigma["A"]))
hook(quote(sigma_B), quote(sigma["B"]))
hook(quote(mu_M), quote(mu["M"]))
hook(quote(zn), quote(z[tau]))
hook(quote(M), quote(overline(X)))
```

@@

```{r, results="asis", echo=TRUE}
f <- function(tau, c, mu_A, sigma_A, mu_B, sigma_B)
{ dfrac(c, mu_A) + (dfrac(1L, mu_A) - dfrac(1L, mu_A + mu_B)) *
    ((mu_A*tau - c) * pnorm(dfrac(c - mu_A*tau, sqrt(sigma_A^2L*tau)))
      - (mu_A*tau + c) * exp(dfrac(2L*mu_A*c, sigma_A^2L))
        * pnorm(dfrac(-c - mu_A*tau, sqrt(sigma_A^2L*tau))))
}

mathout(f)
```

@@

````{r, echo=FALSE}
mu_A = 0.53
mu_B = 1.32
sigma_A = 4.4
sigma_B = 18.3
tau = c(-Inf, -167, -133, -100, -67, -33, 0, 33, 67, 100, 133, 167, Inf)
c = 100
mu_M = 161
````

```{r, results="asis", echo=FALSE}
# Conditions

@@

mathout(g)
```

@@

```{r, echo=FALSE}
mu_A = 0.53
mu_B = 1.32
sigma_A = 4.4
sigma_B = 18.3
tau = -167
c = 100
mu_M = 161

g(tau, c, mu_A, sigma_A, mu_B, sigma_B)
```

@@

```{r, echo=FALSE}
g(tau, c, mu_A, sigma_A, mu_B, sigma_B) + mu_M
```

@@

```{r, results="asis", echo=FALSE}

@@

h <- function(tau, c, mu_A, sigma_A, mu_B, sigma_B, mu_M)
{
  G(tau, c, mu_A, sigma_A, mu_B, sigma_B) + mu_M
}

mathout(h)
```

```{r, echo=FALSE}

@@

h(tau, c, mu_A, sigma_A, mu_B, sigma_B, mu_M)
```

@@

```{r, results="asis", echo=FALSE}
zf <-  function(M, s, N, tau, c, mu_A, sigma_A, mu_B, sigma_B, mu_M)
{
  E <- h(tau, c, mu_A, sigma_A, mu_B, sigma_B, mu_M)
  dfrac(M-E, s/sqrt(N))
}

mathout(zf)
```

```{r, echo=FALSE}
# The vector X contains the observed values, the vector s their observed standard deviation and N = 400 are the number of trials.

@@

zn <- zf(M, s, N, tau, c, mu_A, sigma_A, mu_B, sigma_B, mu_M)
zn
```

@@

```{r, results="asis", echo=FALSE}
par <- c(mu_A = 0.53, sigma_A = 4.4, mu_B = 1.32, sigma_B = 18.3, mu_M = 161)

@@

mathout(gof)
```

```{r, results="asis", echo=FALSE}
n <- gof(par, tau, M, s, N)
n
pchisq(n, 8, lower.tail = FALSE, log.p = FALSE)
```

```{r, results="asis", echo=FALSE}
lsq = function(tau, M, s, N)
{
  start = c(mu_A = 0.3, sigma_A = 2, mu_B = 1 , sigma_B = 5, mu_M = 200)
  optim(start, fn = gof,
    tau = tau, M = M, s = s, N = N, hessian = TRUE, control = list(fnscale = 1))
}

mathout(lsq)

@@

````{r, results="asis", echo=TRUE}
f <- function(tau)
{ dfrac(c, mu["A"]) + (dfrac(1L, mu["A"]) - dfrac(1L, mu["A"] + mu["B"]) * 
    ((mu["A"]*tau - c) * pnorm(dfrac(c - mu["A"]*tau, sqrt(sigma["A"]^2L*tau)))
      - (mu["A"]*tau + c) * exp(dfrac(2L*mu["A"]*tau, sigma["A"]^2L))
        * pnorm(dfrac(-c - mu["A"]*tau, sqrt(sigma["A"]^2L*tau)))))
}

attributes(f)$name = quote(f)
mathout(f)
````

@@

````{r, results="asis", echo=TRUE}
f <- function(tau)
{ dfrac(c, mu["A"]) + (dfrac(1L, mu["A"]) - dfrac(1L, mu["A"] + mu["B"])) * 
    ((mu["A"]*tau - c) * pnorm(dfrac(c - mu["A"]*tau, sqrt(sigma["A"]^2L*tau)))
      - (mu["A"]*tau + c) * exp(dfrac(2L*mu["A"]*tau, sigma["A"]^2L))
        * pnorm(dfrac(-c - mu["A"]*tau, sqrt(sigma["A"]^2L*tau))))
}

attributes(f)$name = quote(f)
mathout(f)
````
